Competing Bandits: The Perils of Exploration under Competition

### 2018-09-03 paper/content/inverted_u.tex
commit updated section 3 and 4

\documentclass[../competing_bandits.tex]{subfiles}
\begin{document}
\section{Competition vs Better Algorithms: Inverted-U}\label{section:5}
Now we demonstrate the inverted-U relationship between the
incentivized algorithms and the competition levels. In particular, we
investigate the incentivized algorithms under three different regimes:
permanent monopoly, temporary monopoly, and permanent duopoly:
\swdelete{We break indifference between exploration algorithms and
  non-exploration algorithms towards non-exploration algorithms due to
  deployment costs.}  \swdelete{We summarize the relationship between
  the different competition levels and incentivized algorithms as
  follows:} \swcomment{think the tie-breaking should be discussed in the prelim?}
\begin{finding}\label{conj:inverted-u}
  \textit{For sufficiently low warm start, the optimal strategies
    \swcomment{for whom??} in the competition game are:
\begin{center}
{Permanent Monopoly} - $\DG$ \\
{Temporary Monopoly} - $\TS$ \\
{Permanent Duopoly} - $\DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.\swcomment{can we quantify?}
}
\end{finding}

\xhdr{Permanent Monopoly.} Since there is only a single firm in the
market for the entire period, the firm can take the entire market
regardless what algorithm it deploys. \gadelete{Assuming the
  deployment of any exploration algorithms will incur certain cost}
\swedit{Since we assume that exploration algorithms ($\DEG$ and $\TS$)
  would incur deployment cost}, the firm would choose to deploy $\DG$.

@@

\xhdr{Permanent Duopoly} \swdelete{Permanent duopoly corresponds to
  the model described in section 2 where $X = 0$ so that both firms
  enter the market simultaneously. What algorithms are firms
  incentivized to deploy? We report the average market share taken
  over the $N$ simulations for a set of exogenous values of
  $T_0$. Additionally, we report the mean and median of the effective
  end of game.}  \swedit{Since both firms enter at the same time, they
  will both get an expected market share of $1/2$ whenever they deploy
  the same algorithms. We will mainly investigate the market shares
  when the firms play different algorithms.  In Tables \ref{sim_ht}
  and \ref{sim_nih}, we report the average market share taken over the
  $N$ simulations for different values of warm-up length $T_0$, and
  also the mean and median of the EEOG, under heavy tail and needle in
  haystack instances.

  For the heavy tail and uniform instances,\footnote{We defer the
    table for uniform instances to the appendix.} we see that $\DG$ is
  a weakly dominant strategy for the competition game, and in
  particular weakly dominates $\TS$. This establishes our claim that
  $\DG$ is the incentivized algorithm.  However, there are two results
  that are at odds with the conjecture: }

\swdelete{Since both firms enter at the same time the game is symmetric and the competition game is zero-sum so we only need to compute the expected payoffs of three pairs of strategies. We do not report them in the figures, but when both firms play the same algorithm the expected market share is 50/50. Since our results are from simulations, we interpret algorithm $A$ being better than algorithm $B$ if the resulting average market share of $A$ is statistically greater than $0.5$. Otherwise, the firm is indifferent between $A$ and $B$ with ties broken towards algorithms that are easier to deploy.

Tables \ref{sim_ht} and \ref{sim_nih} show the results of the simulations. For the most part Conjecture~\ref{conj:inverted-u} is validated where, for the Heavy Tail and Uniform instances, we see that $\DG$ does better than $\TS$ and is a weakly dominant strategy in the game. However, there are two results that are at odds with the conjecture:
}

\swcomment{Guy, please check above! for what instances are $\DG$ a weakly dominant strategy again? Also, what is $N$?}

\begin{puzzle}\label{puz:dg_ht}
Even for large warm start, $\DG$ is better than $\DEG$ under heavy tail instances.\footnote{We demonstrate this holds even for $K = 3$ in the supplementary material}
\end{puzzle}

\begin{puzzle}\label{puz:nih}
Under Needle In Haystack, $\TS$ is weakly dominant across all warm starts. 
\end{puzzle}

\OMIT{
\begin{finding}
\textit{
\begin{enumerate}
\item For low warm starts, under Heavy Tail and Uniform \\ $\DG$ is a weakly dominant strategy in competition (see Table \ref{sim_ht}),
\item  (see Table \ref{sim_nih}). 
\item For sufficiently large warm starts, $\TS$ is dominant across all instances.
\item Even for a large warm start, $\DG$ leads to a higher expected market share than $\DEG$ under Heavy Tail
\end{enumerate}
}
\end{finding}
}

\OMIT{
Looking at the relative reputation plots in Figure \ref{relative_rep_plots}, we can interpret fixing a warm start $T_0$ as fixing the starting point on the relative reputation plots. The proportion of first rounds in the competition game that will go to a firm playing alg $A$ over a firm playing alg $B$ will correspond to the relative reputation proportion at time $k$.
The intuition for the result in competition aligns with the fact that, for Uniform and Heavy Tail, $\TS$ is relative reputation costly compared to $\DG$ and so, for low warm starts, $\TS$ has not yet recovered from the relative reputation loss due to purposeful exploration and so $\DG$ beats $\TS$. However, if we move the warm start sufficiently high so that $\TS$ moves to the relative reputation benefit regime, $\TS > \DG$.
}

 \OMIT{However, we do not always see that $\DG > \TS$ under some warm starts. Table \ref{sim_nih} shows the results for the Needle In Haystack family where we see that $\TS > \DG$. This is precisely due to the fact that, under the Needle In Haystack, $\TS$ is not relative reputation costly compared to $\DG$ and thus we see that $\TS > \DG$.}

We defer the explanation for these puzzles until the next section. Looking at the EEOG values we note that, even though the game continues until $t=2000$, the game effectively ends much earlier than that. This seems to imply that when attempting to understand the performance of different algorithms in competition it is important to look at their performance for relatively small samples instead of asymptotic performance. This is due to the fact that the losing firm may enter into a ``death spiral" where the winner keeps learning and improving from its customers thus increasing its reputation while the losing firm gets starved of customers and progressively falls farther and farther behind its competitors.

@@

\xhdr{Temporary Monopoly.} Recall that the incumbent firm enters the
market and serves as a monopolist until the entrant firm enters at
round $X$. \gaedit{For this section we focus on the warm start values (concretely, $T_0 = 20$) and on the instances where we observed that, under permanent duopoly, $\DG$ was the weakly dominant strategy} \footnote{However, we report the
  results for the Needle In Haystack instances in the supplementary
  material. The results are unsurprising, where $\TS$ is the dominant
  strategy for both the entrant and incumbent.}.

\gadelete{we report results fixing $T_0 = 20$ and
focus on the family of instances where we observed that, for this warm
start, $\DG$ was the optimal strategy.}\swcomment{I don't
  understand this sentence. what's the point?} \swdelete{It seems clear that the temporary monopoly will get a larger market share regardless of what
  algorithm it plays.}

\begin{finding}
  \textit{If the incumbent firm can stay as a temporary monopolist for
    a sufficiently long time, it will deploy the better learning
    algorithm, $\TS$, even though, under the permanent duopoly, they
    would commit to $\DG$.}
\end{finding}

@@

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Aug 15 19:03:20 2018
\begin{table}[ht]
\centering
\caption{Temporary Monopoly Heavy Tail X = 200}
\begin{tabular}{rlll}
\hline
\multicolumn{4}{c}{Incumbent Algorithm}\\
\multirow{12}{0.6in}{\rotatebox{90}{Entrant Algorithm}} \\
  \hline
 & TS & DEG &  DG \\
  \hline
TS & \makecell{\textbf{0.003} $\pm$0.003\\Var:0.002\\ES:100\%} & \makecell{\textbf{0.083} $\pm$0.02\\Var:0.07\\ES:97\%} & \makecell{\textbf{0.17} $\pm$0.02\\Var:0.1\\ES:95\%} \\
  DEG & \makecell{\textbf{0.045} $\pm$0.01\\Var:0.03\\ES:92\%} & \makecell{\textbf{0.25} $\pm$0.02\\Var:0.1\\ES:75\%} & \makecell{\textbf{0.23} $\pm$0.02\\Var:0.1\\ES:78\%} \\
   DG & \makecell{\textbf{0.12} $\pm$0.02\\Var:0.08\\ES:88\%} & \makecell{\textbf{0.36} $\pm$0.03\\Var:0.2\\ES:76\%} & \makecell{\textbf{0.3} $\pm$0.02\\Var:0.1\\ES:64\%} \\
   \hline
\end{tabular}
\label{ht_incum}
\caption*{{The first line in each cell contains the average market share for the entrant over $N=1000$ simulations as well as a 95\% confidence interval. The second line contains the sample variance of the observed market shares and the third line contains the fraction of simulations that ended up with one firm getting an ``extreme" market share of $> 90\%$. Note that smaller values in the table are better for the incumbent. Market shares are calculated as the fraction of users selecting a particular firm \textit{after} the entrant has already entered (i.e. the free rounds to the incumbent do not count towards the share)}}
\end{table}

Why do we observe that $\TS$ is the dominant strategy for the incumbent whereas in the permanent duopoly experiment we saw that, under the same parameters, $\DG$ was preferred to $\TS$? The intuition for this is that competition in the duopoly forces the firms to worry about their reputation which dissuades them from committing to algorithms that involve pure exploration in the early rounds. One can view allowing one firm to temporarily be a monopolist as temporarily relaxing the ``incentive" component of exploration, exploitation, and incentives so that the incumbent firm only faces the classic tradeoff between exploration and exploitation. The incumbent only needs to worry about her reputation after $X$ periods when the entrant comes into the market and again needs to worry about incentivizing agents to select them over their competition.

\OMIT{As a result, the incumbent is incentivized to commit to an algorithm that does exploration in the early rounds since she no longer suffers the same relative reputational cost that she would suffer under competition as long as the $X$ is sufficiently large that she can begin to recover the reputational costs of exploration.} Thus, counterintuitively, by having one firm be a monopoly and dominate the market, we can incentivize them to play $\TS$.

>>>>

\documentclass[../competing_bandits.tex]{subfiles}
\begin{document}
\section{Competition vs Better Algorithms: Inverted-U}\label{section:5}

Now we demonstrate the inverted-U relationship between the
incentivized algorithms and the competition levels. In particular, we
investigate the incentivized algorithms under three different regimes:
permanent monopoly, temporary monopoly, and permanent duopoly: \swcomment{think the tie-breaking should be discussed in the prelim?}

\begin{finding}
\textit{For sufficiently low warm start under the instances that are relative reputation costly, the optimal strategies in the competition game are:
\begin{center}
\textbf{Permanent Monopoly} - $\DG$ \\
\textbf{Temporary Monopoly} (incumbent)- $\TS$ \\
\textbf{Permanent Duopoly} - $\DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.
}
\end{finding}

\xhdr{Permanent Monopoly.} Since there is only a single firm in the
market for the entire period, the firm can take the entire market
regardless of what algorithm it deploys. \gadelete{Assuming the
  deployment of any exploration algorithms will incur certain cost}
\swedit{Since we assume that exploration algorithms ($\DEG$ and $\TS$)
  would incur deployment cost}, the firm would choose to deploy $\DG$.

@@

\xhdr{Permanent Duopoly} \swdelete{Permanent duopoly corresponds to
  the model described in section 2 where $X = 0$ so that both firms
  enter the market simultaneously. What algorithms are firms
  incentivized to deploy? We report the average market share taken
  over the $N$ simulations for a set of exogenous values of
  $T_0$. Additionally, we report the mean and median of the effective
  end of game.}  \swedit{Since both firms enter at the same time, they
  will both get an expected market share of $1/2$ whenever they deploy
  the same algorithms. We will mainly investigate the market shares
  when the firms play different algorithms.  In Tables \ref{sim_ht}
  and \ref{sim_nih}, we report the average market share taken over the
  $N$ simulations for different values of warm-up length $T_0$, and
  also the mean and median of the EEOG, under heavy tail and needle in
  haystack instances.}

For the relative reputation costly instances,\footnote{We defer the table for uniform instances to the appendix.} we see that, for low warm start, $\DG$ is a weakly dominant strategy for the competition game, and in particular weakly dominates $\TS$. This establishes our claim that $\DG$ is the incentivized algorithm.

\gadelete{
\begin{puzzle}\label{puz:dg_ht}
Even for large warm start, $\DG$ is better than $\DEG$ under heavy tail instances.\footnote{We demonstrate this holds even for $K = 3$ in the supplementary material}
\end{puzzle}

\begin{puzzle}\label{puz:nih}
Under Needle In Haystack, $\TS$ is weakly dominant across all warm starts. 
\end{puzzle}
}

The reason we see that $\DG$ weakly dominates under the Heavy Tail and Uniform instances but that $\TS$ is weakly dominant under Needle In Haystack is precisely due to the fact that the purposeful exploration engaged by $\TS$ in the early rounds leads it to suffer a relative reputational cost in the former case but not in the latter. Looking at the relative reputation plots in Figure \ref{relative_rep_plots}, we can interpret fixing a warm start $T_0$ as fixing the starting point on the relative reputation plots. The proportion of first rounds in the competition game that will go to a firm playing alg $A$ over a firm playing alg $B$ will correspond to the relative reputation proportion at time $T_0$. As a result, we see that for warm start $T_0 = 20$ on the relative reputation costly instances, $\DG$ will win the first agent in a larger proportion of the simulations than $\TS$. However, on the Needle In Haystack instances, we see that $\TS$ wins a larger proportion of the first agents compared to $\DG$.

What happens after the first agent? Looking at the EEOG values we note that, even though the game continues until $t=2000$, the game effectively ends much earlier than that. This seems to imply that when attempting to understand the performance of different algorithms in competition it is important to look at their performance for relatively small samples instead of asymptotic performance. This is due to the fact that the losing firm may enter into a ``death spiral" where the winner keeps learning and improving from its customers thus increasing its reputation while the losing firm gets starved of customers and progressively falls farther and farther behind its competitors. As a result, even though if $\TS$ could get sufficiently many agents it would eventually have a higher reputation than $\DG$, since it starts the game by losing agents to $\DG$ due to its lowered reputation from exploration in the warm start it ends up doing worse in the competition game.

@@

\xhdr{Temporary Monopoly.} Recall that the incumbent firm enters the
market and serves as a monopolist until the entrant firm enters at
round $X$. \gaedit{For this section we focus on the warm start values and on the relative reputation costly instances where we observed that, under permanent duopoly, $\DG$ was the weakly dominant strategy. Concretely, our presented results are for $T_0 = 20$.} \footnote{However, we report the results for the Needle In Haystack instances in the supplementary
  material. The results are unsurprising, where $\TS$ is the dominant
  strategy for both the entrant and incumbent.}.

\begin{finding}
  \textit{If the incumbent firm is a temporary monopolist for
    a sufficiently long time, it will deploy the better learning
    algorithm, $\TS$, even though, under the permanent duopoly, they
    would commit to $\DG$.}
\end{finding}

@@

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Aug 15 19:03:20 2018

\begin{table}[ht]
\centering
\caption{Temporary Monopoly Heavy Tail X = 200}
\begin{tabular}{cc|c|c|c|}
  & \multicolumn{1}{c}{} & \multicolumn{3}{c}{Incumbent} \\
  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\TS$}  & \multicolumn{1}{c}{$\DEG$}  & \multicolumn{1}{c}{$\DG$} \\\cline{3-5}
            & $\TS$ & \makecell{\textbf{0.003} \\$\pm$0.003} & \makecell{\textbf{0.083} \\ $\pm$0.02} & \makecell{\textbf{0.17} \\$\pm$0.02} \\ \cline{3-5}
Entrant  & $\DEG$ & \makecell{\textbf{0.045} \\$\pm$0.01} & \makecell{\textbf{0.25} \\$\pm$0.02} & \makecell{\textbf{0.23} \\$\pm$0.02} \\\cline{3-5}
            & $\DG$ & \makecell{\textbf{0.12} \\ $\pm$0.02} & \makecell{\textbf{0.36} \\$\pm$0.03} & \makecell{\textbf{0.3} \\$\pm$0.02} \\\cline{3-5}
\end{tabular}
\label{ht_incum}
\caption*{{The table contains the average market share for the entrant as well as 95\% confidence intervals. Note that smaller values in the table are better for the incumbent. Market shares are calculated as the fraction of users selecting a particular firm \textit{after} the entrant has already entered (i.e. the free rounds to the incumbent do not count towards the share)}}
\end{table}

Why do we observe that $\TS$ is the dominant strategy for the incumbent whereas in the permanent duopoly experiment we saw that, under the same parameters, $\DG$ was the weakly dominant strategy? The intuition for this is that competition in the duopoly forces the firms to worry about their reputation which dissuades them from committing to algorithms that involve pure exploration in the early rounds. One can view allowing one firm to temporarily be a monopolist as temporarily relaxing the ``incentive" component of exploration, exploitation, and incentives so that the incumbent firm only faces the classic tradeoff between exploration and exploitation. The incumbent only needs to worry about her reputation after $X$ periods when the entrant comes into the market and again needs to worry about incentivizing agents to select them over their competition.

As a result, the incumbent is incentivized to commit to an algorithm that does exploration in the early rounds since she no longer suffers the same immediate relative reputational cost that she would suffer under competition as long as the $X$ is sufficiently large that she can begin to recover the reputational costs of exploration. Thus, counterintuitively, by having one firm be a monopoly and dominate the market, we can incentivize them to play $\TS$.

### 2018-09-03 paper/content/perf_in_iso.tex
attempt to make the figures look better
commit updated section 3 and 4

First, we confirm that\gaedit{, for a sufficiently long time horizon,} algorithms' performance is ordered as we'd expect:
    $\TS \gg \DEG \gg \DG$.
For each algorithm and each MAB instance, we compute the mean reputation score at each round, averaged over all \MRVs. We plot the \emph{mean reputation trajectory}: how this score evolves over time. Figure \ref{prelim_means} shows such plot for the Needle In Haystack instance (for other MAB instances the plots are similar). We summarize this finding as follows:

\begin{finding}
\textit{The mean reputation trajectories are arranged as predicted by prior work: 
    $\TS > \DEG > \DG$.}
\end{finding}

\begin{figure}
\caption{Mean reputation trajectories for Needle-in-Haystack. The shaded area displays 95\% confidence intervals.}
\includegraphics[scale=0.35]{"figures/nih_iso_mean"}
\label{prelim_means}
%\caption*{\tiny{Mean reputation trajectory The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of a given algorithm on a given instance and then average this value across all the runs. The shaded area display 95\% confidence intervals.}}
\end{figure}

The mean reputation trajectory is, arguably, the most natural way to represent an algorithm's performance on a given problem instance. But is it the right statistic for our purposes? That is, how well do these trajectories predict the outcomes of the competition game? The following conjecture is natural:

@@

We revisit this conjecture after documenting the results in the competition game.

@@

Let us consider a statistic with a more direct connection to the competition game: for each time $t$, the proportion of \MRVs for which one algorithm has a higher reputation score than the other.

>>>>

First, we confirm that\gaedit{, for a sufficiently long time horizon,} algorithms' performance is ordered as we'd expect:
    $\TS \gg \DEG \gg \DG$.
For each algorithm and each MAB instance, we compute the mean reputation score at each round, averaged over all \MRVs. We plot the \emph{mean reputation trajectory}: how this score evolves over time. Figure \ref{prelim_means} shows such \gaedit{a} plot for the Needle In Haystack instance (for other MAB instances the plots are similar). We summarize this finding as follows:

\begin{finding}
\textit{The mean reputation trajectories\gaedit{, for sufficiently long time horizon,} are arranged as predicted by prior work:
    $\TS > \DEG > \DG$.}
\end{finding}

\begin{figure}
\caption{Mean reputation trajectories for Needle-in-Haystack. The shaded area displays 95\% confidence intervals.}
\includegraphics[scale=0.35]{figures/nih_iso_mean}
\label{prelim_means}
%\caption*{\tiny{Mean reputation trajectory The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of a given algorithm on a given instance and then average this value across all the runs. The shaded area display 95\% confidence intervals.}}
\end{figure}

The mean reputation trajectory is, arguably, the most natural way to represent an algorithm's performance on a given problem instance. But is it the right statistic for our purposes? That is, how well do these trajectories predict the outcomes of the competition game?

Our experiments led us to consider a statistic with a more direct connection to the competition game: for each time $t$, the proportion of \MRVs for which one algorithm has a higher reputation score than the other which we define as the \textit{relative reputation proportion}.

\begin{figure}[ht]
\caption{Relative Reputation Proportion Plots for $\TS$ vs $\DG$. Shaded area display 95\% confidence intervals.}
\includegraphics[scale=0.35]{figures/relative_uniform_annotated_plot}
\includegraphics[scale=0.35]{figures/relative_nih_ts_dg.png}
%\caption*{\tiny{The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of both of the algorithms on a given instance and then calculate the proportion of runs where $\TS \geq \DG$. The shaded area display 95\% confidence intervals.}}
\label{relative_rep_plots}

Figure \ref{relative_rep_plots} shows plots of the relative reputation proportion for $\TS$ vs $\DG$ on the Uniform and Needle In Haystack instances. For the Uniform instance we see that, in the early rounds, the relative reputation proportion for $\TS$  vs. $\DG$ is less than 0.5 meaning that $\DG$ has a higher reputation in a majority of the simulations but that, eventually, $\TS$ has a higher reputation than $\DG$ in a majority of the simulations. The intuition behind this is that, especially since the firms start with no substantive initial information, $\TS$ does purposeful exploration in the early rounds in order to acquire information which leads to a relatively lower reputation compared to $\DG$. However, eventually the information acquired from purposeful exploration in the early rounds allows $\TS$ to make better decisions and achieve a higher reputation, especially when the instance is ``hard enough" so that $\DG$ cannot trivially find the best arm.

\begin{definition}
\textit When purposeful exploration leads to a relative reputation loss compared to the greedy alternative we define this as a \textit{relative reputation cost}. The eventual gain in reputation from the early acquired information leads to what we define as a \textit{relative reputation benefit}. Here we treat it as an empirical definition based on the relative reputation proportion where the ``cost" regime is when the proportion $< 0.5$ and the ``benefit" regime is when the proportion is $> 0.5$ for the better algorithm. We call the instances where there is an early costly period followed by a later benefit period \textit{relative reputation costly instances}.
\end{definition}

\begin{finding}
\textit{Purposeful exploration can lead to relative reputational costs and, as a result, the relative reputation proportion of $\TS$ vs $\DG$ favors $\DG$ for small time horizons. We observe this under the Uniform and Heavy Tail instances. However, it is not observed under the Needle In Haystack instances}
\end{finding}

Is exploration always costly? Figure \ref{relative_rep_plots} also shows that for the Needle In Haystack instances, $\TS$ always does relatively better than $\DG$. There are two contributing factors to this. First, $\TS$ identifies the best arm faster in the Needle In Haystack instances than the Uniform instances so that there is a shorter time horizon where $\TS$ needs to engage in purposeful exploration. Second, in the Needle In Haystack instances there are no ``bad" arms as there may be in the Uniform instances since by construction all the arms except one in Needle In Haystack are the same. Thus, when $\TS$ pulls a sub-optimal arm relative to its current information, the expected reward is the same as the greedy option that has not identified the best arm. However, with the Uniform instances, it is possible that the sub-optimal arm that is pulled has substantially lower expected reward relative to the greedy option. Thus, only the Uniform and Heavy Tail instances are relative reputation costly instances.

\OMIT{The following conjecture is natural:

@@

We revisit this conjecture after documenting the results in the competition game.
}

@@

.
